Graded Call-By-Push-Value

### 2020-07-22 popl21/popl21.tex
[ popl21 ] more typos fixed after submission

%\newcommand{}{\ensuremath{}}
\newcommand{\bla}{\ensuremath{\mbox{$$}}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}\xspace}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

@@

% Concrete effects
\newcommand{\tprint}{\mathsf{print}}
\newcommand{\print}[2]{\tprint\,#1.\,#2}
\newcommand{\toutput}{\mathsf{output}}
%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
\newcommand{\tthrow}{\mathsf{throw}}
\newcommand{\throw}[1]{\tthrow\,#1}
\newcommand{\tcatch}{\mathsf{catch}}
\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
\newcommand{\traise}{\mathsf{raise}}
%\newcommand{\raise}[1]{\traise\,#1} %clash
\newcommand{\thandle}{\mathsf{handle}}
\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}

@@

Recall that effects are modeled by a suitable monad $\T : \CC \to \CC$
in $\CC$.  We shall refer to elements of a monadic type $\T\,A$ as
\emph{computations} when $\CC = \SET$.  For a general $\CC$,
computations shall be the morphisms $\Hom A {\T\,B}$ called
\emph{Kleisli arrows}.
An example module would be the \emph{writer monad}
$\T\,A = \String \times A$
that models the effect \emph{output}.
Another example would be the \emph{exception monad} $\Exc + \_$ where
the result of a computation is either an exception $e : \Exc$ or a
regular result.
The monad operations, together with
their implementation for \emph{writer} in $\SET$, are the following:
\[
\begin{array}{lcl}
\fmap[\T] & : & (A \to B) \to \T\,A \to \T\,B \\
\fmap[\T] f\, (s,\; a) & = & (s,\, f\, b)
\\[1ex]
\return[\T] & : & A \to \T\,A \\
\return[\T]\, a & = & (\verb|""|, a)
\\[1ex]
\join[\T] & : & \T\,(\T\,A) \to \T\,A \\
\join[\T]\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a) \\
\end{array}
\]
Herein, \verb|""| shall denote the empty string and $\append$ string
concatenation.

@@

The presence of $\tfmap : \Hom A B \to \Hom {\T\,A} {\T\,B}$
satisfying the functor laws $\tfmap\,\tid = \tid$ and
$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$ makes $\T$ a
\emph{(endo)functor}, written $\T : [\CC \to \CC]$.  The
endofunctors $F : [\CC \to \CC]$ form a category, the
$\emph{functor category}$, with identity $\Id\,A = A$ and
composition $(F \comp G)\,A = F\,(G\,A)$.  We shall write
$f : F \todot G$ for morphisms in the functor category, called
\emph{natural transformations}.  These are families
$f_A : \Hom {F\,A} {G\,A}$ of morphisms that commute with the
functor action $\tfmap$, \ie,
$\fmap[G]h \comp f_A = f_B \comp \fmap[F]h$ for any $h : \Hom A B$.

Unit $\treturn : \Id \todot \T$ and multiplication
$\tjoin : \T \comp \T \todot \T$ are \emph{natural transformations}
which breaks down to $\fmap[\T] h \comp \treturn = \treturn \comp h$
and $\fmap[\T] h \comp \tjoin = \tjoin \comp \fmap{(\fmap h)}$.
%
The three equational
laws of $\tjoin$ can be visualized compactly in the following
commutative diagram.
\[
\xymatrix@C=12ex{
  \T\,A     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[d]^{\tjoin}
& \T\,(\T\,(\T\,A)) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\tjoin}
\\
& \T\,A       \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[l]_{\tjoin}
\\
&
& \T\,A \ar[u]_{\tfmap\,\treturn}
}
\]

@@

Recall that a \emph{monoidal category} $\EE$ has a designated object
$\I : \EE$ and an operation $\_{\otimes}\_ : [\EE \to [\EE \to \EE]$,
the \emph{tensor product} on objects of $\EE$ that is functorial in
both positions.  Further, there are natural isomorphisms
$\Gl : (\I \otimes A) \cong A$ and $\Gr : (A \otimes \I) \cong A$
witnessing the unitality of $\I$ and
$\alpha : (A \otimes (B \otimes C)) \cong ((A \otimes B) \otimes C)$
witnessing associativity of $\otimes$.

@@

Call-by-push value \cite{levy:hosc06} is a refinement of Moggi's
computational lambda-calculus \cite{moggi:infcomp91} that allows effects
not only in monadic types, \ie, in objects $\T\,A$, but more generally
in computation types.  These correspond to monad algebras for $\T$ in $\CC$,
aka $\T$-algebras.
Those algebras are objects $B$ together with a morphism $\run[B] : \T\,B
\to B$ that allows to formally \emph{run} the monad, ``merging'' its
effects into $B$.
The prime example of a monad algebra is simply a monadic
type, because $\run[\T\,A] : \T\,(\T\,A) \to \T\,A$ is just
$\tjoin$.  Levy \citeyearpar{levy:hosc06} shows that monad algebras are
closed under products ($\times$) and exponentiation ($\To$)
with arbitrary objects.  \Eg,
in $\SET$ we can define, for \emph{writer} algebras:
\[
\begin{array}{lcl}
%   \run[B_1 \times B_2] & : & \T\,(B_1 \times B_2) \to B_1 \times B_2 \\
%   \run[B_1 \times B_2]\, (s,(b_1,b_2)) & = & (\run[B_1] (s,b_1),\
%                                            \run[B_2] (s,b_2))
% \\[1ex]
  \run[B \times B'] & : & \T\,(B \times B') \to B \times B' \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T\,(A \To B) \to (A \To B) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]
These definition implement \emph{lazy} effects that cannot be observed
at computation types such as $A \To B$, but only at value types; the
$\run[]$ of the monad algebra pushes the effects towards result types
that are eventually types of observable objects (values).

@@

Graded monad algebras are as well closed under
pointwise products $(B \times B')_e = B_e \times B'_e$ and
exponentiation with objects $(A \To B)_e = A \to B_e$;
here the graded writer example:
\[
\begin{array}{lcl}
  \run[B \times B'] & : & \T[e_1]\,(B \times B')_{e_2} \to (B \times
                          B')_{e_1 \bu e_2} \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T[e_1]\,(A \To B)_{e_2} \to (A \To B)_{e_1 \bu e_2} \\
  % \run[A \To B] & : & \T[e_1]\,(A \To B_{e_2}) \to (A \To B_{e_1 \bu e_2}) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]

\begin{remark}[Naturality of $\trun$]
  The $\T$-algebra $B : [\Eff \to \CC]$ comes with a natural morphism
  $\run[e_1,e_2] : \Hom {\T[e_1]\,B_{e_2}} {B_{e_1 \bu e_2}}$ in the
  sense that
  $B_{(e_1\leq e_1') \bu (e_2 \leq e_2')} \comp {\run[e_1,e_2]} =
  {\run[e_1',e_2']} \comp \T[e_1 \leq e_1']\,B_{e_2 \leq e_2'}$.
\end{remark}

@@

Types are classified into value types $P \in \PTy$ (written $A$ in
\citet{levy:hosc06}) and computation types $N \in \NTy$
(written $\underline B$ in
\loccit).  These are positive and negative types in the terminology of
focusing \cite{zeilberger:PhD}.
Positive types are generated from base types $o$ via
disjoint sums $\sumty i I P$ with tag set $I$, eager products $\tupty
i I P$ of arity $I$ (composed
from $1$ and $A_1 \times A_2$ in \loccit) and thunking $\thunkty e N$
(written $U\,\underline B$ in \loccit).  In contrast to pure CBPV,
thunk types $\thunkty e N$ are annotated with an effect $e$ that can
be triggered when the thunk is forced.  Negative types are just as in
pure CBPV: monadic types $\diamond P$ (written $F\,A$ in \loccit),
function types $P \To N$ (written $A \to \underline B$) in \loccit)
and record types $\recty i I N$ with label set $I$.
Records are \emph{lazy} tuples whose components are only computed by demand.
We abbreviate $\sumty i I P$ by its ``meta-level $\eta$-contraction''
$\sumt I P$; notations $\tupt I P$ and
$\rect I N$ are understood analogously.

Terms are separated into values $v \in \PTm$ and computations $t \in
\NTm$ and identical to pure CBPV, modulo changes in the concrete
syntax.  Values introduce positive types,
computations introduce
negative types and
eliminate both positive and negative types.
We use bars to indicate sequences, e.g., $\bar v$ for a
sequence of values, but drop the bar when the context of discourse
makes clear that we are dealing with sequences rather than single
objects.  For instance, ``$v_i$'' (where $i : I$) in the premise of
\rintro\otimes indicates that $v$ is a sequence of values with
elements $(v_i)_{i:I}$.  We may abbreviate $\trecord \{i :
t_i\}_{i:I}$ by $\recrd I t$ where $I$ is the label set and $t$ a
mapping from labels $i:I$ to terms $t_i$.

@@

The denotational semantics of the novel parts
of graded CBPV has been informally explained in
sections \ref{sec:ccc} to \ref{sec:graded_algebra} already; in the following,
we spell out the details.  We assume a distributive cartesian-closed
category $\CC$ with a strong graded monad $\T : [\Eff \to [\CC \to \CC]]$.
Let us agree on some notation for the constructions on objects and
morphisms:
\begin{itemize}
\item Product $\prod_{i:I} A_i$ with projections $\pi_i : \CC(\prod_I
  A, A_i)$ and tupling $\langle f_i
  \rangle_{i:I} : \CC(C,\prod_i A)$ for $f_i : \CC(C,A_i)$.  Binary
  products $\prod_{\{1,2\}} A$ are written $A_1 \times A_2$,
  and the nullary product (terminal object)
  is written $1$ with nullary tupling $\langle\rangle : \CC(C,1)$.
\item Coproduct $\coprod_{i:I} A_i$ with injections $\iota_i :
  \CC(A_i,\coprod_I A)$ and cotupling $[f_i]_{i:I} : \CC(\coprod_I A,
  B)$ for $f_i : \CC(A_i,B)$.
\item Exponential $A \To B$ with $\Lambda : \CC(C \times A, B) \to
  \CC(C, A \To B)$ and $\teval : \CC((A \To B) \times A, B)$.
\item Graded monad $\T$ with functoriality $\fmap[\T] f :
  \CC(\T[e]A,\T[e]B)$ for $f : \CC(A,B)$,
  unit $\treturn : \CC(A,\T[\Ge]A)$,
  multiplication $\tjoin : \CC(\T[e_1](\T[e_2]A),\T[e_1 \bu e_2]A)$,
  strength $\strengthl : \CC(A \times \T[e]B, \T[e](A \times B))$ and
  coercion $\tcast : \CC(\T[e]A,\T[e']A)$ for $e \leq e'$.
  Costrength $\strengthr : \CC(\T[e]A \times B, \T[e](A \times B))$
  is derivable in the standard way.
\end{itemize}

@@

Given a set $\Exc$ of exceptions whose elements we refer to by $e$ and whose subsets by $E$,
consider the effect algebra $(\Pot\Exc,\cup,\emptyset,\subseteq)$.
Primitives for throwing and catching exceptions can be added to graded
CBPV by the following rules:
\newcommand{\eee}{E_1\setminus\{e\} \cup E_2}
\[
  \ru{}{\CompTy \Gamma {\throw e} {\{e\}} N} %{e : \Exc}
\qquad
  \ru{\CompTy \Gamma u {E_1} N \qquad
      \CompTy \Gamma t {E_2} N
    }{\CompTy \Gamma {\catch u e t} {\eee} N}
\]
If $u$ throws exception $e$, then ``$\catch u e t$'' computes $t$, else $u$.
%
In $\SET$ we use the graded monad $\T[E]\,A = E + A$ and a family of
morphism $\traise\,e : \SET(1,\T[\{e\}]A)$ to interpret
$\dent{\throw e} = \traise\,e \comp \langle\rangle$.  To interpret
$\tcatch$, define a family
$\thandle_N\,e : \Hom[\SET] {N_{E_1} \times N_{E_2}} {N_{\eee}}$ by
induction on $N \in \NTy$:
\[
\begin{array}{lcl}
  \handle[\diamond P] e g h & = & \left\{
                                  \begin{array}{ll}
                                    h & \mbox{if } g = \iota_1 e \\
                                    g & \mbox{otherwise}
                                  \end{array}
\right. \\
  \handle[P \To N] e g h \, a & = & \handle[N] e {(g\,a)} {(h\,a)} \\
  \handle[\tupt I N] e g h \, i & = & \handle[N_i] e {(g\,i)} {(h\,i)} \\
\end{array}
\]
Catching is then
$\dent {\catch u e t} = \handle[N] e \comp \langle \dent u, \dent t
\rangle$.
Note that $\thandle$ cannot be defined in terms of $\trun$, but we
have to break $N$ down to monadic type $\diamond P$
to get access to the exception
thrown by $u$.

@@

Not all sequences of effects are always meaningful: For instance,
reading from a file before it was opened is impossible and could be
prevented statically by graded effect typing.  This could be modelled
by adding a maximal element $\top : \Eff$---with $e \leq \top$ for all
$e : \Eff$ that signifies an inconsistent state.  This error element
would also be dominant in sequences, \ie,
$\top \bu e = e \bu \top = \top$.
A program $\CompTy \Gamma t e N$ would only be accepted if $e \not= \top$.

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
%\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tdisplay}{\mathsf{display}}
\newcommand{\display}[1][]{\tdisplay_{#1}}
% \newcommand{\tduplicate}{\mathsf{display}}
% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\mathsf{head}}
\newcommand{\ttail}{\mathsf{tail}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}
\newcommand{\tiso}{\mathsf{iso}}
\newcommand{\isot}{\tiso^\otimes}

@@

Let us consider the \emph{stream comonad}%
\footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
$\D\,B = \bN \to B$,
with $\thead\,s = s\,0$ and $\ttail\,s = s \comp \suc$.
The stream comonad lets us work in a setting where values are not
single data points, but streams of data.
Besides functoriality, a
comonad has the natural transformations $\textract$ and $\display$
(in category theory called $\Ge$ and $\Gd$),
dual to $\treturn$ and $\tjoin$.  The implementation of the stream
comonad in $\SET$ is as follows:
\[
\begin{array}{lcl}
  \textract & : & \D\,B \to B \\
  \textract\,s & = & \thead\,s
\\[1.5ex]
  \tdisplay & : & \D\,B \to \D\,(\D\,B) \\
  \thead\,(\tdisplay\,s) & = & s \\
  \ttail\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
\end{array}
\]
The generic operations of a comonad allow us to \emph{extract} the
value wrapped in a comonadic structure and to \emph{display} another
layer of that structure.

>>>>

%\newcommand{}{\ensuremath{}}
\newcommand{\bla}{\ensuremath{\mbox{$$}}}
\newcommand{\ie}{\emph{i.e.}\xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\Eg}{\emph{E.g.}\xspace}
\newcommand{\loccit}{\emph{loc.\,cit.}\xspace}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

@@

% Concrete effects
\newcommand{\tprint}{\mathsf{print}}
\newcommand{\print}[2]{\tprint\,#1.\,#2}
\newcommand{\toutput}{\mathsf{output}}
%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
\newcommand{\tthrow}{\mathsf{throw}}
\newcommand{\throw}[1]{\tthrow\,#1}
\newcommand{\tcatch}{\mathsf{catch}}
\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
\newcommand{\traise}{\mathsf{raise}}
%\newcommand{\raise}[1]{\traise\,#1} %clash
\newcommand{\thandle}[1][]{\mathsf{handle}_{#1}}
\newcommand{\handle}[4][]{\thandle[{#1}]\,#2\,(#3,#4)}

@@

Recall that effects are modeled by a suitable monad $\T : \CC \to \CC$
in $\CC$.  We shall refer to elements of a monadic type $\T\,A$ as
\emph{computations} when $\CC = \SET$.  For a general $\CC$,
computations shall be the morphisms $\Hom A {\T\,B}$ called
\emph{Kleisli arrows}.
An example monad would be the \emph{writer monad}
$\T\,A = \String \times A$
that models the effect \emph{output}.
Another example would be the \emph{exception monad} $\Exc + \_$ where
the result of a computation is either an exception $e : \Exc$ or a
regular result.
The monad operations, together with
their implementation for \emph{writer} in $\SET$, are the following:
\[
\begin{array}{lcl}
\fmap[\T] & : & (A \to B) \to \T\,A \to \T\,B \\
\fmap[\T] f\, (s,\; a) & = & (s,\, f\, b)
\\[1ex]
\return[\T] & : & A \to \T\,A \\
\return[\T]\, a & = & (\verb|""|, a)
\\[1ex]
\join[\T] & : & \T\,(\T\,A) \to \T\,A \\
\join[\T]\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a) \\
\end{array}
\]
Herein, \verb|""| shall denote the empty string and $\append$ string
concatenation.

@@

The presence of $\tfmap : \Hom A B \to \Hom {\T\,A} {\T\,B}$
satisfying the functor laws $\tfmap\,\tid = \tid$ and
$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$ makes $\T$ an
\emph{(endo)functor}, written $\T : [\CC \to \CC]$.  The
endofunctors $F : [\CC \to \CC]$ form a category, the
$\emph{functor category}$, with identity $\Id\,A = A$ and
composition $(F \comp G)\,A = F\,(G\,A)$.  We shall write
$f : F \todot G$ for morphisms in the functor category, called
\emph{natural transformations}.  These are families
$f_A : \Hom {F\,A} {G\,A}$ of morphisms that commute with the
functor action $\tfmap$, \ie,
$\fmap[G]h \comp f_A = f_B \comp \fmap[F]h$ for any $h : \Hom A B$.

Unit $\treturn : \Id \todot \T$ and multiplication
$\tjoin : \T \comp \T \todot \T$ are \emph{natural transformations}---%
which breaks down to $\fmap[\T] h \comp \treturn = \treturn \comp h$
and $\fmap[\T] h \comp \tjoin = \tjoin \comp \fmap{(\fmap h)}$.
%
The three equational
laws of $\tjoin$ can be visualized compactly in the following
commutative diagram.
\[
\xymatrix@C=12ex{
  \T\,A     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[d]^{\tjoin}
& \T\,(\T\,(\T\,A)) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\tjoin}
\\
& \T\,A       \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[l]_{\tjoin}
\\
&
& \T\,A \ar[u]_{\tfmap\,\treturn}
}
\]

@@

Recall that a \emph{monoidal category} $\EE$ has a designated object
$\I : \EE$ and an operation $\_{\otimes}\_ : [\EE \to [\EE \to \EE]]$,
the \emph{tensor product} on objects of $\EE$ that is functorial in
both positions.  Further, there are natural isomorphisms
$\Gl : (\I \otimes A) \cong A$ and $\Gr : (A \otimes \I) \cong A$
witnessing the unitality of $\I$ and
$\alpha : (A \otimes (B \otimes C)) \cong ((A \otimes B) \otimes C)$
witnessing associativity of $\otimes$.

@@

Call-by-push value \cite{levy:hosc06} is a refinement of Moggi's
computational lambda-calculus \citeyearpar{moggi:infcomp91} that allows effects
not only in monadic types, \ie, in objects $\T\,A$, but more generally
in computation types.  These correspond to monad algebras for $\T$ in $\CC$,
aka $\T$-algebras.
Those algebras are objects $B$ together with a morphism $\run[B] : \T\,B
\to B$ that allows to formally \emph{run} the monad, ``merging'' its
effects into $B$.
The prime example of a monad algebra is simply a monadic
type, because $\run[\T\,A] : \T\,(\T\,A) \to \T\,A$ is just
$\tjoin$.  Levy \citeyearpar{levy:hosc06} shows that monad algebras are
closed under products ($\times$) and exponentiation ($\To$)
with arbitrary objects.  \Eg,
in $\SET$ we can define, for \emph{writer} algebras:
\[
\begin{array}{lcl}
%   \run[B_1 \times B_2] & : & \T\,(B_1 \times B_2) \to B_1 \times B_2 \\
%   \run[B_1 \times B_2]\, (s,(b_1,b_2)) & = & (\run[B_1] (s,b_1),\
%                                            \run[B_2] (s,b_2))
% \\[1ex]
  \run[B \times B'] & : & \T\,(B \times B') \to B \times B' \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T\,(A \To B) \to (A \To B) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]
These definition implement \emph{lazy} effects that cannot be observed
at computation types such as $A \To B$, but only at value types; the
$\run[]$ of the monad algebra pushes the effects towards result types
that are eventually types of observable objects (values).

@@

Graded monad algebras are, like non-graded ones, closed under
pointwise products $(B \times B')_e = B_e \times B'_e$ and
exponentiation with objects $(A \To B)_e = A \to B_e$;
here the graded writer example:
\[
\begin{array}{lcl}
  \run[B \times B'] & : & \T[e_1]\,(B \times B')_{e_2} \to (B \times
                          B')_{e_1 \bu e_2} \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T[e_1]\,(A \To B)_{e_2} \to (A \To B)_{e_1 \bu e_2} \\
  % \run[A \To B] & : & \T[e_1]\,(A \To B_{e_2}) \to (A \To B_{e_1 \bu e_2}) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]

\begin{remark}[Naturality of $\trun$]
  The $\T$-algebra $B : [\Eff \to \CC]$ comes with a \emph{natural} morphism
  $\run[e_1,e_2] : \Hom {\T[e_1]\,B_{e_2}} {B_{e_1 \bu e_2}}$ in the
  sense that
  $B_{(e_1\leq e_1') \bu (e_2 \leq e_2')} \comp {\run[e_1,e_2]} =
  {\run[e_1',e_2']} \comp \T[e_1 \leq e_1']\,B_{e_2 \leq e_2'}$.
\end{remark}

@@

Types are classified into value types $P \in \PTy$ (written $A$ in
\citet{levy:hosc06}) and computation types $N \in \NTy$
(written $\underline B$ in
\loccit).  These are positive and negative types in the terminology of
focusing \cite{zeilberger:PhD}.
Positive types are generated from base types $o$ via
disjoint sums $\sumty i I P$ with tag set $I$,
eager products $\tupty i I P$
(composed from $1$ and $A_1 \times A_2$ in \loccit)
of arity $I$
and thunking $\thunkty e N$
(written $U\,\underline B$ in \loccit).  In contrast to pure CBPV,
thunk types $\thunkty e N$ are annotated with an effect $e$ that can
be triggered when the thunk is forced.  Negative types are just as in
pure CBPV: monadic types $\diamond P$ (written $F\,A$ in \loccit),
function types $P \To N$ (written $A \to \underline B$) in \loccit)
and record types $\recty i I N$ with label set $I$.
Records are \emph{lazy} tuples whose components are only computed by demand.
We abbreviate $\sumty i I P$ by its ``meta-level $\eta$-contraction''
$\sumt I P$; notations $\tupt I P$ and
$\rect I N$ are understood analogously.

Terms are separated into values $v \in \PTm$ and computations $t \in
\NTm$ and are identical to pure CBPV, modulo changes in the concrete
syntax.  Values introduce positive types,
computations introduce
negative types and
eliminate both positive and negative types.
We use bars to indicate sequences, e.g., $\bar v$ for a
sequence of values, but drop the bar when the context of discourse
makes clear that we are dealing with sequences rather than single
objects.  For instance, ``$v_i$'' (where $i : I$) in the premise of
\rintro\otimes indicates that $v$ is a sequence of values with
elements $(v_i)_{i:I}$.  We may abbreviate $\trecord \{i :
t_i\}_{i:I}$ by $\recrd I t$ where $I$ is the label set and $t$ a
mapping from labels $i:I$ to terms $t_i$.

@@

The denotational semantics of the novel parts
of graded CBPV has been informally explained in
sections \ref{sec:ccc} to \ref{sec:graded_algebra} already; in the following,
we spell out the details.  We assume a distributive cartesian-closed
category $\CC$ with a strong graded monad $\T : [\Eff \to [\CC \to \CC]]$.
Let us agree on some notation for the constructions on objects and
morphisms:
\begin{itemize}
\item Product $\prod_{i:I} A_i$ with projections $\pi_i : \CC(\prod_I
  A, A_i)$ and tupling $\langle f_i
  \rangle_{i:I} : \CC(C,\prod_i A)$ for $f_i : \CC(C,A_i)$.  Binary
  products $\prod_{\{1,2\}} A$ are written $A_1 \times A_2$,
  and the nullary product (terminal object)
  is written $1$ with nullary tupling $\langle\rangle : \CC(C,1)$.
\item Coproduct $\coprod_{i:I} A_i$ with injections $\iota_i :
  \CC(A_i,\coprod_I A)$,
  cotupling $[f_i]_{i:I} : \CC(\coprod_I A, B)$ for $f_i : \CC(A_i,B)$,
  and distribution morphism
  $\tdist : \Hom {C \times \coprod_I A} {\coprod_{i:I} {(C \times A_i)}}$.
\item Exponential $A \To B$ with $\Lambda : \CC(C \times A, B) \to
  \CC(C, A \To B)$ and $\teval : \CC((A \To B) \times A, B)$.
\item Graded monad $\T$ with functoriality $\fmap[\T] f :
  \CC(\T[e]A,\T[e]B)$ for $f : \CC(A,B)$,
  unit $\treturn : \CC(A,\T[\Ge]A)$,
  multiplication $\tjoin : \CC(\T[e_1](\T[e_2]A),\T[e_1 \bu e_2]A)$,
  strength $\strengthl : \CC(A \times \T[e]B, \T[e](A \times B))$ and
  coercion $\tcast : \CC(\T[e]A,\T[e']A)$ for $e \leq e'$.
  Costrength $\strengthr : \CC(\T[e]A \times B, \T[e](A \times B))$
  is derivable in the standard way.
\end{itemize}

@@

Given a set $\Exc$ of exceptions whose elements we refer to by $e$ and whose subsets by $E$,
consider the effect algebra $(\Pot\Exc,\cup,\emptyset,\subseteq)$.
Primitives for throwing and catching exceptions can be added to graded
CBPV by the following rules:
\newcommand{\eee}{E_1\setminus\{e\} \cup E_2}
\[
  \ru{}{\CompTy \Gamma {\throw e} {\{e\}} N} %{e : \Exc}
\qquad
  \ru{\CompTy \Gamma u {E_1} N \qquad
      \CompTy \Gamma t {E_2} N
    }{\CompTy \Gamma {\catch u e t} {\eee} N}
\]
If $u$ throws exception $e$, then ``$\catch u e t$'' computes $t$, else $u$.
%
In $\SET$ we use the graded monad $\T[E]\,A = E + A$ and a family of
morphism $\traise\,e : \SET(1,\T[\{e\}]A)$ to interpret
$\dent{\throw e} = \traise\,e \comp \langle\rangle$.  To interpret
$\tcatch$, define a family
$\thandle[N]\,e : \Hom[\SET] {N_{E_1} \times N_{E_2}} {N_{\eee}}$ by
induction on $N \in \NTy$:
\[
\begin{array}{l@{~}lcl}
  \handle[\diamond P] {&e} g h & = & \left\{
                                  \begin{array}{ll}
                                    h & \mbox{if } g = \iota_1 e \\
                                    g & \mbox{otherwise}
                                  \end{array}
\right. \\
  \handle[P \To N] {&e} g h \, a & = & \handle[N] e {g\,a} {h\,a} \\
  \handle[\tupt I N] {&e} g h \, i & = & \handle[N_i] e {g\,i} {h\,i} \\
\end{array}
\]
Catching is then
$\dent {\catch u e t} = \thandle[N]\,e \comp \langle \dent u, \dent t
\rangle$.
Note that $\thandle$ cannot be defined in terms of $\trun$, but we
have to break $N$ down to monadic type $\diamond P$
to get access to the exception
thrown by $u$.

@@

Not all sequences of effects are always meaningful: For instance,
reading from a file before it was opened is impossible and could be
prevented statically by graded effect typing.  This could be modelled
by adding a maximal element $\top : \Eff$ (with $e \leq \top$ for all
$e : \Eff$) that signifies an inconsistent state.  This error element
would also be dominant in sequences, \ie,
$\top \bu e = e \bu \top = \top$.
A program $\CompTy \Gamma t e N$ would only be accepted if $e \not= \top$.

@@

\newcommand{\lolli}{\multimap}
\newcommand{\qfun}[3]{#1#2 \lolli #3}
\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
%\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
\newcommand{\nbox}[1]{}
\newcommand{\tzip}{\mathsf{zip}}
\newcommand{\textract}{\mathsf{extract}}
\newcommand{\extract}[1][]{\textract_{#1}}
\newcommand{\tdisplay}{\mathsf{display}}
\newcommand{\display}[1][]{\tdisplay_{#1}}
% \newcommand{\tduplicate}{\mathsf{display}}
% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
\newcommand{\tduplicate}{\mathsf{duplicate}}
\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
\newcommand{\tcobind}{\mathsf{cobind}}
\newcommand{\tbind}{\mathsf{bind}}
\newcommand{\tget}{\mathsf{get}}
\newcommand{\tmodify}{\mathsf{modify}}
\newcommand{\thead}{\ensuremath{\mathsf{head}}}
\newcommand{\ttail}{\ensuremath{\mathsf{tail}}}
\newcommand{\suc}{(\_{+}1)}
\newcommand{\Float}{\mathsf{Float}}
\newcommand{\tiso}{\mathsf{iso}}
\newcommand{\isot}{\tiso^\otimes}

@@

Let us consider the \emph{stream comonad}%
\footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
$\D\,B = \bN \to B$,
with $\thead\,s = s\,0$ and $\ttail\,s = s \comp \suc$.
The stream comonad lets us work in a setting where values are not
single data points, but streams of data.
Besides functoriality, a
comonad has the natural transformations $\textract$ and $\display$
(in category theory called $\Ge$ and $\Gd$),
dual to $\treturn$ and $\tjoin$.  The implementation of the stream
comonad in $\SET$ is as follows:
\[
\begin{array}{lcl}
  \textract & : & \D\,B \to B \\
  \textract\,s & = & \thead\,s
\\[1.5ex]
  \tdisplay & : & \D\,B \to \D\,(\D\,B) \\
  \thead\,(\tdisplay\,s) & = & s \\
  \makebox[0pt][l]{\ttail}\phantom{\thead}\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
\end{array}
\]
The generic operations of a comonad allow us to \emph{extract} the
value wrapped in a comonadic structure and to \emph{display} another
layer of that structure.
